\pkg{pksensi}: An R package to apply sensitivity analysis in pharmacokinetic models

### 2019-02-24 manuscript/manuscript.Rmd
update Introduction

Sensitivity analysis is a mathematical technique to investigate how variations in model parameters affect model outputs. An increasing number of studies use sensitivity analysis to determine which model parameters contribute to high variation in model predictions \citep{ferretti2016trends}. Also, this technique has been applied in pharmacology and toxicology research \citep{loizou2015application, mcnally2012reconstruction}. Pharmacokinetic modeling describes the changes in the concentrations or amounts of a substance within model compartments over time. The goal of sensitivity analysis in pharmacokinetic research is to examine the sensitivity of output variables (e.g. compound concentration in blood or tissues) in pharmacokinetic models responds to input parameters, such as anatomical, physiological, and kinetic constants \citep{mcnally2011workflow}. It can be further applied to parameter prioritization and parameter fixing before model calibration \citep{fphar201800588}.

@@

Most of sensitivity analysis tools are constructed using scientific computing software, such as \proglang{GNU Octave}/\proglang{Matlab} \citep{pianosi2015matlab}, \proglang{python} \citep{herman2017salib}, and \proglang{R} \citep{iooss2017introduction}. Previously, we mainly used \proglang{R} with the \pkg{sensitivity} package \citep{R-sensitivity}, which is a practical tool to conduct local and global sensitivity analysis. The sensitivity package includes some functions to generate the parameter sequences by using different computing algorithms. Also, it provides a feasible way to integrate external modeling results. This computational approach can effectively solve the heavy computing burden of numerical solutions within the pure \proglang{R} environment. Several \proglang{R} packages, such as \pkg{FME}, \pkg{multisensi}, and \pkg{ODEsensitivity} include sensitivity analysis tools and are freely available in the Comprehensive R Archive Network (CRAN). The \pkg{FME} package contains a basic method for performing local sensitivity analysis for dynamic models \citep{JSSv033i03}. The \pkg{multisensi} package provides sensitivity analysis for models with multivariate output \citep{R-multisensi}. \pkg{ODEsensitivity} can perform sensitivity analysis in ordinary differential equation (ODE) models \citep{R-ODEsensitivity}. It uses the ODE interface from the deSolve package, connecting it with the sensitivity analysis from the sensitivity package, and can perform sensitivity analysis on models with multivariate output. Other related packages such as \pkg{mtk} \citep{RJ-2015-031} and \pkg{fast} \citep{R-fast} are designed to invetigate the influence of model parameter. Although these tools provide various approaches to perform sensitivity analysis, we have not find a suitable package that provides functions with which users can visualize and distinguish the “non-influential” model parameters and can further apply to parameter fixing in pharmacokinetic modeling.  
Here we develop an \proglang{R} package, called \pkg{pksensi}, which is designed to make sensitivity analysis more accessible and reproducible in pharmacological and toxicological researches. This package can investigate both parameter uncertainty and sensitivity in pharmacokinetic models, such as PBPK, and advanced compartment absorption and transit models with multivariate model output. The design concepts of \pkg{pksensi} are:

>>>>

Sensitivity analysis is a mathematical technique to investigate how variations in model parameters affect model outputs. An increasing number of studies use sensitivity analysis to determine which model parameters contribute to high variation in model predictions \citep{ferretti2016trends}. This technique has also been applied in pharmacology and toxicology research \citep{loizou2015application, mcnally2012reconstruction}. Pharmacokinetic modeling describes the changes in the concentrations or amounts of a substance within model compartments over time. The goal of sensitivity analysis in pharmacokinetic research is to examine the sensitivity of output variables (e.g. compound concentration in blood or tissues) in pharmacokinetic models responds to input parameters, such as anatomical, physiological, and kinetic constants \citep{mcnally2011workflow}. It can be further applied to parameter prioritization and parameter fixing before model calibration \citep{fphar201800588}.

@@

Most sensitivity analysis tools are constructed using scientific computing software, such as \proglang{GNU Octave}/\proglang{Matlab} \citep{pianosi2015matlab}, \proglang{python} \citep{herman2017salib}, and \proglang{R} \citep{iooss2017introduction}. Previously, we mainly used \proglang{R} with the \pkg{sensitivity} package \citep{R-sensitivity}, which is a practical tool to conduct local and global sensitivity analysis. The sensitivity package includes some functions to generate the parameter sequences by using different computing algorithms. Also, it provides a feasible way to integrate external modeling results. This computational approach can effectively solve the heavy computing burden of numerical solutions within the pure \proglang{R} environment. Several other \proglang{R} packages, such as \pkg{FME}, \pkg{multisensi}, and \pkg{ODEsensitivity} include sensitivity analysis tools and are freely available in the Comprehensive R Archive Network (CRAN). The \pkg{FME} package contains a basic method for performing local sensitivity analysis for dynamic models \citep{JSSv033i03}. The \pkg{multisensi} package provides sensitivity analysis for models with multivariate output \citep{R-multisensi}. \pkg{ODEsensitivity} can perform sensitivity analysis in ordinary differential equation (ODE) models \citep{R-ODEsensitivity}. It uses the ODE interface from the deSolve package, connecting it with the sensitivity analysis from the sensitivity package, and can perform sensitivity analysis on models with multivariate output. Other related packages such as \pkg{mtk} \citep{RJ-2015-031} and \pkg{fast} \citep{R-fast} are designed to investigate the influence of model parameter. Although these tools provide various approaches to perform sensitivity analysis, we have not found a suitable package that provides functions with which users can visualize and distinguish the “non-influential” model parameters and can further apply to parameter fixing in pharmacokinetic modeling.  
Wepresent here an \proglang{R} package, called \pkg{pksensi}, which is designed to make sensitivity analysis more accessible and reproducible in pharmacological and toxicological researches. This package can investigate both parameter uncertainty and sensitivity in pharmacokinetic models, such as PBPK, and advanced compartment absorption and transit models with multivariate model output. The design concepts of \pkg{pksensi} are:
